Online Multiclass Linear Classification with Bandit Feedback

### 2019-01-15 ICML-submission/algorithm-for-strongly-linearly-separable-data.tex
finished section 4

To finish the proof, we need to upper bound the number of updates $U$. We claim
that $U \le \lfloor 4(R/\gamma)^2 \rfloor$. The proof of this upper bound is
similar to the proof of the mistake bound for \textsc{Multiclass Perceptron}
algorithm. Let $w_1^*, w_2^*, \dots, w_K^* \in V$ be vectors that satisfy
\eqref{equation:strong-linear-separability-1},
\eqref{equation:strong-linear-separability-2} and
\eqref{equation:strong-linear-separability-3}.
The $K$-tuple $(w_1^{(t)}, w_2^{(t)}, \dots, w_K^{(t)})$
changes only if there is an update in round $t$.
We investigate how $\sum_{i=1}^K \norm{w_i^{(t)}}^2$ and
$\sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}}$ change. If there is an update in round $t$,
by lines~\ref{line:pos-update} and~\ref{line:neg-update}, we always have
$ w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} + (-1)^{z_t} x_t $,
and for all $i \neq \widehat y_t$, $w_{i}^{(t+1)} = w_{i}^{(t)}$.
Therefore,
\begingroup
\allowdisplaybreaks
\begin{align*}
& \sum_{i=1}^K \norm{w_i^{(t+1)}}^2 \\
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}} \norm{w_i^{(t)}}^2 \right) + \norm{w_{\widehat y_t}^{(t+1)}}^2 \\
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}} \norm{w_i^{(t)}}^2 \right) + \norm{w_{\widehat y_t}^{(t)} + (-1)^{z_t} x_t}^2 \\
& = \left( \sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + \norm{x_t}^2 + \underbrace{(-1)^{z_t} 2 \ip{w_{\widehat y_t}^{(t)}}{x_t}}_{\le 0} \\
& \le \left( \sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + \norm{x_t}^2 \\
& \le \left( \sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + R^2 \; .
\end{align*}
\endgroup
Hence, after $U$ updates,
\begin{equation}
\sum_{i=1}^K \norm{w_i^{(T+1)}}^2 \le R^2 U \; .
\label{eqn:norm-ub}
\end{equation}
Similarly, if there is an update in round $t$, we have
\begingroup
\allowdisplaybreaks
\begin{align*}
& \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \\
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}} \ip{w_i^*}{w_i^{(t)}} \right) + \ip{w_{\widehat y_t}^*}{w_{\widehat y_t}^{(t+1)}} \\
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}} \ip{w_i^*}{w_i^{(t)}} \right) \\
& \qquad + \ip{w_{\widehat y_t}^*}{w_{\widehat y_t}^{(t)} + (-1)^{z_t} x_t} \\
& = \left( \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \right) + (-1)^{z_t} \ip{w_{\widehat y_t}^*}{x_t} \\
& \ge \left( \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \right) + \frac \gamma 2,
\end{align*}
\endgroup
where the last inequality follows from a case analysis on $z_t$ and
Definition~\ref{definition:strong-linear-separability}: if $z_t = 0$, then
$\widehat y_t = y_t$, by Equation~\eqref{equation:strong-linear-separability-2},
we have that $\ip{w_{\widehat y_t}^*}{x_t} \geq \frac \gamma 2$; if $z_t = 1$,
then $\widehat y_t \neq y_t$, by
Equation~\eqref{equation:strong-linear-separability-3}, we have that
$\ip{w_{\widehat y_t}^*}{x_t} \le -\frac \gamma 2$.

@@

The upper bound $(K-1) \lfloor 4(R/\gamma)^2 \rfloor$ on the expected number of
mistakes of
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples} is
optimal to within constant factor as long as the number of classes $K$ is
at most $O((R/\gamma)^2)$. This is proved in
\autoref{theorem:strongly-separable-examples-mistake-lower-bound} below.

@@

\paragraph{Remark.} Irrespective of any conditions on $K$, $R$, and $\gamma$, a trivial lower bound
on the expected number of mistakes of any randomized algorithm is
$\frac{K-1}{2}$. (Adversary chooses the same example over and over.) However, it
is unclear if the trivial lower bound is the best possible if $K$ is large,
i.e., $\omega((R/\gamma)^2)$. We leave this as an open problem.

@@

for $i \in \{1,2,\dots,K\}$ and any $j \in
\{1,2,\dots,M\}$, we consider the inner product of $w_i^*$ and $v_j$.
If $i = \ell_j$, $\ip{w_i^*}{v_j} = \gamma - \frac \gamma 2 = \frac \gamma 2$;
otherwise $i \neq \ell_j$, in this case,
$\ip{w_i^*}{v_j} = 0 - \frac \gamma 2 = - \frac \gamma 2$.
This means that $w_1^*, w_2^*, \dots, w_K^*$ satisfy
conditions
\eqref{equation:strong-linear-separability-2} and
\eqref{equation:strong-linear-separability-3}. Condition \eqref{equation:strong-linear-separability-1}
is satisfied since
\begin{multline*}
\sum_{i=1}^K \norm{w_i^*}^2
= 2 \frac{\gamma^2}{R^2} \sum_{j=1}^M \norm{e_j}^2 +  \frac{\gamma^2}{2 R^2} K \norm{e_{M+1}}^2 \\
= 2 \frac{\gamma^2}{R^2} M + \frac{\gamma^2}{2 R^2} K
\le \frac{1}{2} + \frac{1}{2}
= 1 \; .
\end{multline*}

>>>>

To finish the proof, we need to upper bound the number of updates $U$. We claim
that $U \le \lfloor 4(R/\gamma)^2 \rfloor$ with probability 1.
The proof of this upper bound is
similar to the proof of the mistake bound for \textsc{Multiclass Perceptron}
algorithm. Let $w_1^*, w_2^*, \dots, w_K^* \in V$ be vectors that satisfy
\eqref{equation:strong-linear-separability-1},
\eqref{equation:strong-linear-separability-2} and
\eqref{equation:strong-linear-separability-3}.
The $K$-tuple $(w_1^{(t)}, w_2^{(t)}, \dots, w_K^{(t)})$
changes only if there is an update in round $t$.
We investigate how $\sum_{i=1}^K \norm{w_i^{(t)}}^2$ and
$\sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}}$ change. If there is an update in round $t$,
by lines~\ref{line:pos-update} and~\ref{line:neg-update}, we always have
$ w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} + (-1)^{z_t} x_t $,
and for all $i \neq \widehat y_t$, $w_{i}^{(t+1)} = w_{i}^{(t)}$.
Therefore,
\begingroup
\allowdisplaybreaks
\begin{align*}
& \sum_{i=1}^K \norm{w_i^{(t+1)}}^2 \\
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}} \norm{w_i^{(t)}}^2 \right) + \norm{w_{\widehat y_t}^{(t+1)}}^2 \\
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}} \norm{w_i^{(t)}}^2 \right) + \norm{w_{\widehat y_t}^{(t)} + (-1)^{z_t} x_t}^2 \\
& = \left( \sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + \norm{x_t}^2 + \underbrace{(-1)^{z_t} 2 \ip{w_{\widehat y_t}^{(t)}}{x_t}}_{\le 0} \\
& \le \left( \sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + \norm{x_t}^2 \\
& \le \left( \sum_{i=1}^K \norm{w_i^{(t)}}^2 \right) + R^2 \; .
\end{align*}
\endgroup
The inequality that $(-1)^{z_t} 2 \ip{w_{\widehat y_t}^{(t)}}{x_t} \leq 0$ is from a case analysis: if line~\ref{line:pos-update} is executed, then $z_t = 0$ and $\ip{w_i^{(t)}}{x_t} \ge 0$; otherwise line~\ref{line:neg-update} is executed, in which case $z_t = 1$ and $\ip{w_i^{(t)}}{x_t} \le 0$.

Hence, after $U$ updates,
\begin{equation}
\sum_{i=1}^K \norm{w_i^{(T+1)}}^2 \le R^2 U \; .
\label{eqn:norm-ub}
\end{equation}
Similarly, if there is an update in round $t$, we have
\begingroup
\allowdisplaybreaks
\begin{align*}
& \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \\
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}} \ip{w_i^*}{w_i^{(t)}} \right) + \ip{w_{\widehat y_t}^*}{w_{\widehat y_t}^{(t+1)}} \\
& = \left( \sum_{i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}} \ip{w_i^*}{w_i^{(t)}} \right) \\
& \qquad + \ip{w_{\widehat y_t}^*}{w_{\widehat y_t}^{(t)} + (-1)^{z_t} x_t} \\
& = \left( \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \right) + (-1)^{z_t} \ip{w_{\widehat y_t}^*}{x_t} \\
& \ge \left( \sum_{i=1}^K \ip{w_i^*}{w_i^{(t)}} \right) + \frac \gamma 2,
\end{align*}
\endgroup
where the last inequality follows from a case analysis on $z_t$ and
Definition~\ref{definition:strong-linear-separability}: if $z_t = 0$, then
$\widehat y_t = y_t$, by Equation~\eqref{equation:strong-linear-separability-2},
we have that $\ip{w_{\widehat y_t}^*}{x_t} \geq \frac \gamma 2$; if $z_t = 1$,
then $\widehat y_t \neq y_t$, by
Equation~\eqref{equation:strong-linear-separability-3}, we have that
$\ip{w_{\widehat y_t}^*}{x_t} \le -\frac \gamma 2$.

@@

The upper bound $(K-1) \lfloor 4(R/\gamma)^2 \rfloor$ on the expected number of
mistakes of
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples} is
optimal upto constant factor, as long as the number of classes $K$ is
at most $O((R/\gamma)^2)$. This is proved in
\autoref{theorem:strongly-separable-examples-mistake-lower-bound} below.

@@

\paragraph{Remark.} Irrespective of any conditions on $K$, $R$, and $\gamma$, a trivial lower bound
on the expected number of mistakes of any randomized algorithm is
$\frac{K-1}{2}$. To see this, note that the adversary can choose an example $(R e_1, y)$,
where $y$ is a label chosen uniformly from $\cbr{1,2\ldots,K}$, and show this example $K-1$ times.
The sequence of examples trivially satisfies the strong linear separability condition, and
the $\frac{K-1}{2}$ expected mistake lower bound follows from Claim 2 of~\citet{Daniely-Helbertal-2013}.
However, it is unclear if the trivial lower bound is the best possible if $K$ is large,
i.e., $\omega((R/\gamma)^2)$. We leave this as an open problem.

@@

for $i \in \{1,2,\dots,K\}$ and $j \in
\{1,2,\dots,M\}$, we consider the inner product of $w_i^*$ and $v_j$.
If $i = \ell_j$, $\ip{w_i^*}{v_j} = \gamma - \frac \gamma 2 = \frac \gamma 2$;
otherwise $i \neq \ell_j$, in which case 
$\ip{w_i^*}{v_j} = 0 - \frac \gamma 2 = - \frac \gamma 2$.
This means that $w_1^*, w_2^*, \dots, w_K^*$ satisfy
conditions
\eqref{equation:strong-linear-separability-2} and
\eqref{equation:strong-linear-separability-3}. Condition \eqref{equation:strong-linear-separability-1}
is satisfied since
\begin{multline*}
\sum_{i=1}^K \norm{w_i^*}^2
= 2 \frac{\gamma^2}{R^2} \sum_{j=1}^M \norm{e_j}^2 +  \frac{\gamma^2}{2 R^2} K \norm{e_{M+1}}^2 \\
= 2 \frac{\gamma^2}{R^2} M + \frac{\gamma^2}{2 R^2} K
\le \frac{1}{2} + \frac{1}{2}
= 1 \; .
\end{multline*}
